Spending too much time at the Galactic bar: chaotic fanning of the Ophiuchus stream

### 2016-01-25 ms.tex
remove old comments

%
%\subsection{Detecting and characterizing chaos with frequency diffusion}\label{sec:chaos-indicator}
%
%Regular orbits are quasi-periodic: any phase-space component of the orbit can be decomposed in terms of a Fourier series, e.g.,
%\begin{equation}
%	x(t) = \sum^\infty_k \, a_{x,k}\, e^{i\, \omega_k \, t}.
%\end{equation}
%The amplitudes, $\bs{a}_k$ are constants which determine the strength of each mode, and the frequencies in each Fourier mode, $\omega_k$, are linear, integer combinations of the fundamental frequencies of the orbit, $\bs{\Omega}$---that is,
%\begin{equation}
%	\omega_k = \bs{n}_k \cdot \bs{\Omega}
%\end{equation}
%where $\bs{n}_k$ is a integer vector with dimensionality equal to the number of degrees of freedom of the gravitational potential \citep[e.g.,][]{binney82, binneytremaine}. The fundamental frequencies of a regular orbit are directly related to the orbital actions, a set of integrals of motion intricately related to the [...] of Hamiltonian and nonlinear dynamics \citep[e.g.,][]{??}. The fundamental frequencies can be approximated for numerically-integrated orbital time series by measuring the positions and amplitudes of spectral lines in the Fourier transform of the orbital components. A very accurate procedure for numerically computing the frequencies was developed by \citet{laskar93} and has been re-implemented in \python\ and released as open-source software \citep[\superfreq;][]{superfreq}. 
%
%Chaotic orbits have no actions and therefore no fundamental frequencies. The frequencies and actions drift erratically in time with rates that depend on the structure of resonances surrounding such orbits. The evolution of the frequencies changes the configuration-space appearance of finite sub-sections of chaotic orbits and greatly enhances the density evolution of ensembles of nearby chaotic orbits; this has a dramatic effect on the morphological evolution of tidal streams \citep[e.g.,][]{apw15-chaos}. Numerically-integrated trajectories can be identified as chaotic orbits by the detection of frequency diffusion \citep[e.g.,][]{laskar96, valluri98}. 
%
%The accuracy of the frequency-finding method (\superfreq) allows for detecting very small changes in the frequencies of even mildly chaotic orbits. Measuring the rate of frequency diffusion is therefore a powerful tool for detecting and characterizing dynamical chaos \citep[e.g.,][]{many}. From measuring the frequencies in many consecutive, overlapping time-windows along a numerically integrated orbit, we define the frequency diffusion rate, $DERP$, as
%\begin{equation}
%	yo
%\end{equation}
%where the index $k$ refers to the window number and $\tau$ is the shift in time between two consecutive windows. We note that this definition differs from previous, coarse approximations of the diffusion rate that simply differenced the frequencies computed in two consecutive time windows. We use this frequency diffusion rate to identify chaotic orbits integrated in the potential models described above.

@@

\begin{table*}[ht]
\footnotesize
\begin{center}
	\begin{tabular}{cccccc}
	\toprule
%	name & $\phi_2$ [deg] & $d$ [kpc] & $\mu_l$ [mas yr$^{-1}$] & $\mu_b$ [mas yr$^{-1}$] & $v_r$ [km s$^{-1}$] & $s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
%static & $-0.03\pm0.05$ & $8.35\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.1\pm0.9$ & $0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
%bar1--9 & $-0.03\pm0.05$ & $8.40\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$ & $0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.9$\\
	name & $\phi_2$ [deg] & $d$ [kpc] & $\mu_l$ [mas yr$^{-1}$] & $\mu_b$ [mas yr$^{-1}$] & $v_r$ [km s$^{-1}$]\\\midrule
	static & $-0.03\pm0.05$ & $8.3\pm0.05$ & $-7.4\pm0.1$ & $0.9\pm0.1$ & $288.9\pm0.9$\\
	bar1--9 & $-0.03\pm0.05$ & $8.35\pm0.05$ & $-7.4\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$\\
	\bottomrule
	\end{tabular}

	\begin{tabular}{ccc}
	\toprule
%	name & $\phi_2$ [deg] & $d$ [kpc] & $\mu_l$ [mas yr$^{-1}$] & $\mu_b$ [mas yr$^{-1}$] & $v_r$ [km s$^{-1}$] & $s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
%static & $-0.03\pm0.05$ & $8.35\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.1\pm0.9$ & $0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
%bar1--9 & $-0.03\pm0.05$ & $8.40\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$ & $0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.9$\\
	$s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
	$0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
	$0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.9$\\
	\bottomrule
	\end{tabular}
	\caption{Estimated mean and standard deviation of samples from the marginal posterior distributions over each parameter in our orbit fit model (the posterior distributions are very close to Gaussian). For the barred potentials, all mean values are the same because the time-dependence of the bar doesn't impact the orbit fit over the short length of the stream. We have made samples from the full posterior distribution available with this article and provide code to transform to and from stream coordinates (see http://adrian.pw/ophiuchus/coordinates for more information).\label{tbl:param-means} }
\end{center}
\end{table*}

@@

%We have applied the same distance-modulus cut to the points in these figures as used to select the BHB stars identified as stream members by S16 (see Section~2 in S16).

@@

The model streams presented here do not reproduce all observed features of the Ophiuchus stream (the details of the potential and the orbit predict vastly different phase-space morphologies for the fanned part). Instead, these results illustrate that chaotic evolution of tidal debris can plausibly explain the peculiar features of the stream. If the cluster progenitor was on a regular orbit, it would have to have disrupted entirely within the last $\approx$300--600 Myr in order to explain the shortness and density profile. In addition, the four most recently identified BHB stars with similar distances and line-of-sight velocities would have to be (highly unlikely) chance alignments of halo stars. If instead the progenitor were on a chaotic orbit (because of the influence of the Galactic bar): (1) stars stripped early will have ``fanned'' out and would thus be harder to observe and (2) the nearby, high-velocity-dispersion BHB stars can be naturally explained by this chaotic stream-fanning. We consider the second scenario to be more plausible: our understanding of the formation and evolution of the Ophiuchus stream is that the progenitor object has been orbiting and steadily losing stars over the last several Gigayears, but only the stars stripped from the most recent few pericentric passages remain coherent enough to be detected as a stream-like over-density in the PS1 data.
%[Gaia, Don't do distance cut that Brani did, chemistry]

@@

%\todo{Where to put this?}
%
%also, cite gnedin97
%
%A typical inner MW globular cluster with a mass $m_{\rm cl} \approx 10^4~\msun$ has a half-mass radius $r_h \approx 2~{\rm pc}$. We can ask, given an assumed mass density profile for the inner galaxy, what Galactocentric pericentric radius the Ophiuchus progenitor would have to reach such that the tidal radius of the cluster is equal to the half-mass radius.
%
%Let's assume the density profile of the inner $\approx 1~{\rm kpc}$ of the MW is a Jaffe spheroid so that
%\begin{align}
%	\rho(r) &= \frac{\rho_0}{(r/a)^2 \, (1 + r/a)^2}\\
%	M(<r) &= 4\pi \, \rho_0 \, a^3 \, \frac{r/a}{1 + r/a}
%\end{align}
%with scale radius $a \approx 0.5~{\rm kpc}$ and mass within the scale radius $M(<a) \approx 10^{10}~\msun$ ($\rho_0 \approx \frac{10^{10}~\msun}{2\pi a^3} \approx 10^{10}~\msun~{\rm kpc}^{-3}$). The tidal radius is approximately
%\begin{align}
%	r_{\rm tide} \approx \left(\frac{m_{\rm cl}}{3M(<r)}\right)^{1/3} \, r
%\end{align}
%so that 
%
%then
%\begin{align}
%	2~{\rm pc} &= \left(\frac{m_{\rm cl}}{3M(<r)}\right)^{1/3} \, r\\
%\end{align}
%
%
%if we set
%\begin{align}
%	r_{\rm tide} = r_h = 2~{\rm pc} 
%\end{align}

>>>>

@@

\begin{table*}[ht]
\footnotesize
\begin{center}
	\begin{tabular}{cccccc}
	\toprule
	name & $\phi_2$ [deg] & $d$ [kpc] & $\mu_l$ [mas yr$^{-1}$] & $\mu_b$ [mas yr$^{-1}$] & $v_r$ [km s$^{-1}$]\\\midrule
	static & $-0.03\pm0.05$ & $8.3\pm0.05$ & $-7.4\pm0.1$ & $0.9\pm0.1$ & $288.9\pm0.9$\\
	bar1--9 & $-0.03\pm0.05$ & $8.35\pm0.05$ & $-7.4\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$\\
	\bottomrule
	\end{tabular}

	\begin{tabular}{ccc}
	\toprule
	$s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
	$0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
	$0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.9$\\
	\bottomrule
	\end{tabular}
	\caption{Estimated mean and standard deviation of samples from the marginal posterior distributions over each parameter in our orbit fit model (the posterior distributions are very close to Gaussian). For the barred potentials, all mean values are the same because the time-dependence of the bar doesn't impact the orbit fit over the short length of the stream. We have made samples from the full posterior distribution available with this article and provide code to transform to and from stream coordinates (see http://adrian.pw/ophiuchus/coordinates for more information).\label{tbl:param-means} }
\end{center}
\end{table*}

@@

@@

The model streams presented here do not reproduce all observed features of the Ophiuchus stream (the details of the potential and the orbit predict vastly different phase-space morphologies for the fanned part). Instead, these results illustrate that chaotic evolution of tidal debris can plausibly explain the peculiar features of the stream. If the cluster progenitor was on a regular orbit, it would have to have disrupted entirely within the last $\approx$300--600 Myr in order to explain the shortness and density profile. In addition, the four most recently identified BHB stars with similar distances and line-of-sight velocities would have to be (highly unlikely) chance alignments of halo stars. If instead the progenitor were on a chaotic orbit (because of the influence of the Galactic bar): (1) stars stripped early will have ``fanned'' out and would thus be harder to observe and (2) the nearby, high-velocity-dispersion BHB stars can be naturally explained by this chaotic stream-fanning. We consider the second scenario to be more plausible: our understanding of the formation and evolution of the Ophiuchus stream is that the progenitor object has been orbiting and steadily losing stars over the last several Gigayears, but only the stars stripped from the most recent few pericentric passages remain coherent enough to be detected as a stream-like over-density in the PS1 data.

@@
